Quantifying stochastic noise in cultured\\ circadian reporter cells

### 2015-07-04 response_letter.tex

Please see our detailed responses to the reviewers below:

\section*{Reviewer \#1}
\begin{reviewer}
The revision improved the manuscript, but a couple of minor issues remain to be addressed.

The added discussion on types of noise should be revised to include the work of Elowitz et al (Science 2002) on extrinsic vs intrinsic noise. The Elowitz study does quantitatively distinguish extrinsic vs intrinsic noise in their experiments, while the revised manuscript states this is only possible in theory (so the wording should be altered at the bottom of page 6 and the Elowitz study cited). 
\end{reviewer}

We thank the reviewer for recommending this important study, and agree that further discussion of existing work on biological noise is useful in the introduction. We have therefore added an additional point to the paragraph on intrinsic and extrinsic noise sources:

\begin{manuscript}{Page 3}
Extrinsic noise results from heterogeneity between cells, such as differences in size or physical environment, leading to differences on a cell-to-cell basis in expected period and amplitude. 
{\bfseries The effects of noise in biological systems has been well-studied, and relative amounts of intrinsic and extrinsic noise can be identified through carefully designed single-cell experiments (15). }
For circadian systems, intrinsic noise has been {\bfseries suspected} to play a larger role: a single cell's variability in period from cycle-to-cycle is larger than the variability in mean period length between cells (2).
\end{manuscript}

For the discussion on page 6 where we describe what can be inferred from experimental data, we had restricted our claims to population-level recordings - whereas the Elowitz study used data from single-cell imaging to differentiate between noise sources. We nevertheless acknowledge the potential for misunderstanding and have added further text describing experimental techniques which might differentiate between noise sources (with a further citation to the work done by Elowitz, {\itshape et al}):

\begin{manuscript}{Pages 6-7}
It is also possible that damping rate changes due to siRNA or small molecule perturbation could be manifested through altering the system's extrinsic noise. 
Such a change, caused perhaps by an unequal uptake of drug across cells, would result increased cell heterogeneity, and different dephasing kinetics. 
While differentiating between intrinsic and extrinsic noise sources from population-level data is possible in theory (Figure S6), these differences are likely not identifiable from typical data (Figure S7). 
{\bfseries This distinction would be possible with single-cell level data, as has been done in other experimental systems (15). 
However, such an experiment would likely not be amenable to high-throughput methods.}
Differences in damping rates are therefore best viewed as representative of changes to overall stochastic noise from both intrinsic and extrinsic factors.
\end{manuscript}

@@

\begin{reviewer}
Also, in the revised part of the introduction addressing the role of noise in characterizing oscillations, the work by Bratsun et al (PNAS 2005) should be mentioned as another caution. For instance, they point out that ``the 'noise signature' of a system dominated by delay can be similar to a system dominated by intrinsic noise. This result implies that care must taken in attributing variability to purely stochastic sources, because delay-induced variability can appear empirically similar to fluctuations arising from intrinsic noise.''
\end{reviewer}

In Bratsun {\itshape et al, PNAS} 2005, the authors demonstrate that variability in gene transcription can occur both through intrinsic noise as well as through fast, deterministic oscillations. While

@@

\begin{reviewer}
In the new expanded paragraph discussing the effect of cell period heterogeneity, the authors dismiss the significance of intrinsic differences in period across the population compared to cycle-to-cycle fluctuations and don't directly quantify the relative contributions. However, Cohen et al (Journal Theoretical Biology 2013) quantified it for this same set of fibroblasts using a Bayesian hierarchical model and found the standard deviation of periods across the population to be 0.89h compared to the within-cell variability 1.43h. So while cycle-to-cycle fluctuations are indeed larger, the variability across the population is still quite substantial, and these values should be mentioned in this section to quantify the relative contributions of these sources of variability to the overall decay rate.
\end{reviewer}

Indeed, we agree with the reviewer that adding quantitative measures adds to the discussion of the relative importance of intrinsic vs.\ extrinsic period variability. We have therefore added these values from both studies. In addition, we have softened some of the language to avoid dismissing the effect of extrinsic variability, which likely indeed plays a non-negligible role in determining the damping rate.

\begin{manuscript}{Page 6}
  While it is true that the dephasing of a group of oscillators can be caused by both variance in the mean period as well as cycle-to-cycle variability, intrinsic stochastic noise {\bf may} play a more significant role. 
{\bf We} show that there is greater variance in period on a cycle-to-cycle basis than on a cell-to-cell basis in cultured fibroblast cells {\bf (Figure S5): individual cell period lengths have an average inner quartile range (IQR) of 3.18 h, while cell-to-cell average period has an IQR of 1.55 h. 
These results are independently confirmed by a previous study using Bayesian modeling, which found a standard deviation of periods within cells of 1.43 h, and 0.89 h across the population (27). 
A similar} result has also been observed in dispersed SCN neurons, suggesting that while both intrinsic and extrinsic period heterogeneity likely contribute to the dephasing kinetics, cell-to-cell differences are less severe than cell-autonomous noise (2).

### 2015-07-05 response_letter.tex

\begin{manuscript}{Pages 6-7}
It is also possible that damping rate changes due to siRNA or small molecule perturbation could be manifested through altering the system's extrinsic noise. 
Such a change, caused perhaps by an unequal uptake of drug across cells, would result increased cell heterogeneity, and different dephasing kinetics. 
While differentiating between intrinsic and extrinsic noise sources from population-level data is possible in theory (Figure S6), these differences are likely not identifiable from typical data (Figure S7). 
{\bfseries This distinction would be possible with single-cell level data, as has been done in other experimental systems (15). 
However, such an experiment would likely not be amenable to high-throughput methods.}
Differences in damping rates are therefore best viewed as representative of changes to overall stochastic noise from both intrinsic and extrinsic factors.
\end{manuscript}

\begin{reviewer}
Also, in the revised part of the introduction addressing the role of noise in characterizing oscillations, the work by Bratsun et al (PNAS 2005) should be mentioned as another caution. For instance, they point out that ``the 'noise signature' of a system dominated by delay can be similar to a system dominated by intrinsic noise. This result implies that care must taken in attributing variability to purely stochastic sources, because delay-induced variability can appear empirically similar to fluctuations arising from intrinsic noise.''
\end{reviewer}

In Bratsun {\itshape et al, PNAS} 2005, the authors demonstrate that variability in the amplitude of gene transcription can occur both through intrinsic noise as well as through fast, deterministic oscillations. While it might be possible that fast oscillations in gene transcription (in addition to the slower, 24-hour circadian rhythms) underly most of the variation seen in ultradian frequencies, it would then be unlikely that we would observe the random drift in phases. Since our study deals mainly with the phase precision rather than the variability in transcription, we have decided to omit this reference to avoid confusing the reader.

@@

{\bf It is also} possible that damping rate changes due to siRNA or small molecule perturbation could be manifested through altering the system's extrinsic noise.

\end{manuscript}

\section*{Reviewer \#4}

\begin{reviewer}
In short, the authors didn't really address the concern about applying their simple model for deductive reasoning. 
They claim that as the siRNA method is well-accepted by the circadian community that their interpretation should be legit. 
Biological response is proportional to dose and target. 
It seems implicitly that the authors are assuming that the DOSE of siRNA transfected into every individual cell is exactly the same. 
The biological response is then interpreted solely as being attributed to the target. 
One look at a flow cytometry result of a siRNA experiment will tell you that the dose is not the same between cells. 
Different siRNA also don't work equally well either. 
So again the criticism that the approach is not convincing in providing the claimed insight still is unaddressed.
\end{reviewer}

We agree with the reviewer that the effect of an siRNA on each individual cell will indeed be varied.
This effect would be manifested in changing the extrinsic noise of the system, as cell heterogeneity would be amplified by the distribution of siRNA effect.
While we previously mentioned variability in drug uptake as a possible source of cell heterogeneity, we have expanded this section to make the reference to siRNA more explicit:

\begin{manuscript}{Pages 6-7}
It is also possible that damping rate changes due to siRNA or small molecule perturbation could be manifested through altering the system’s extrinsic noise. 
{\bfseries Such a change could be caused by an unequal uptake of siRNA or drug on a cell-to-cell basis, as has been demonstrated by a distribution of single-cell knockdown efficiency through flow cytometry (27). 
This effect would increase cell heterogeneity and lead to faster dephasing kinetics.}
\end{manuscript}

While it is also true that different siRNA constructs will have different knockdown effectivenesses, this will be a systematic effect that will not effect either the intrinsic or extrinsic noise.
In developing our approach, we do not assume that each siRNA will be equally effective - just as the knockdown of different genes will have varied effects on circadian rhythms.
While we do group siRNA perturbations by target gene, this test mainly serves to guard against off-target activities.
We have added a relevant explanation to the methods section clarifying this point:

\begin{manuscript}{Page 10}
A Hotelling's $T^2$ test was used to determine whether the means of each gene knockdown was significantly different from the control population.
While different siRNA constructs will have different knockdown efficiencies, grouping based on gene target helps to eliminate the effect of off-target activities. 
\end{manuscript}

Once more, we would like to thank each of our reviewers for their time and attention to detail.
We believe their follow-up critiques have further improved the rigor of this work, and have included our appreciation for their efforts in the acknowledgements.

\vspace{4ex}
\begin{flushright}
  Sincerely,\\[2ex]

  \includegraphics[width=0.3\textwidth]{/home/peter/Documents/Research/figures/frank_signature.pdf}\\[1ex]
Francis J. Doyle III
\end{flushright}

\end{document}

\end{document}

### 2015-07-05 main_text.tex
adding discussion of Elowitz et al
finishing reviewer updates

In the preceding sections, we have demonstrated that changes to single-cell intrinsic noise are sufficient to explain the observed changes in population-level damping.
However, experimental work has shown that cell-autonomous fibroblast cells have a distribution of mean free-running periods \cite{Leise2012}.
Indeed, prior to the availability of single-cell data, studies explored the possibility that differences in mean periods served as the mechanism behind population-level damping \cite{Izumo2003}.
While it is true that the dephasing of a group of oscillators can be caused by both variance in the mean period as well as cycle-to-cycle variability, intrinsic stochastic noise may play a more significant role. 
We show that there is greater variance in period on a cycle-to-cycle basis than on a cell-to-cell basis in cultured fibroblast cells (Figure~\ref{fig:cell2cell-variability}): individual cell period lengths have an average inner quartile range (IQR) of 3.18 h, while cell-to-cell average period has an IQR of 1.55 h. 
These results are independently confirmed by a previous study using Bayesian modeling, which found a standard deviation of periods within cells of 1.43 h, and 0.89 h across the population \cite{Cohen2012}.
A similar result has also been observed in dispersed SCN neurons, suggesting that while both intrinsic and extrinsic period heterogeneity likely contribute to the dephasing kinetics, cell-to-cell differences are less severe than cell-autonomous noise \cite{Herzog2004}.

It is also possible that damping rate changes due to siRNA or small molecule perturbation could be manifested through altering the system's extrinsic noise.
Such a change, caused perhaps by an unequal uptake of drug across cells, would result increased cell heterogeneity, and different dephasing kinetics.
While differentiating between intrinsic and extrinsic noise sources from population-level data is possible in theory (Figure~\ref{fig:dephasing-rates}), these differences are likely not identifiable from typical data (Figure~\ref{fig:computational_dephasing}).
Differences in damping rates are therefore best viewed as representative of changes to overall stochastic noise from both intrinsic and extrinsic factors.
Since both types of noise are important to determining the overall function of population-level rhythms, damping rates are still a valuable method of quantifying stochastic noise.

@@

In this study, we have described a method by which changes to stochastic noise can be estimated from population-level circadian bioluminescence recordings.
An overview of the computational steps involved in this method are outlined in Figure~\ref{fig:comp_overview}.
While the method can be applied to existing experimental data, there are also practical considerations for the design of future experiments.
Because the damping rate must be inferred from the time-varying amplitude, collecting bioluminescence data for longer time periods yields more accurate results and reduces the potential impact of initial transient regions.
Additionally, a sampling rate that is high enough to confidently capture the peaks and troughs of gene expression is required -- in this study, the 2 hr sampling window of the siRNA screen proved sufficient.
While achieving such a rate is typically not difficult for bioluminescence experiments, it may limit the method's applicability in experiments where samples need to be analyzed at each time point.
We also note that there are many available tools for detrending and regressing time-series data.
While we prioritized computationally efficient methods (which could be scaled to genome-wide screens), the best methods for any particular application may vary depending on the data.

@@

\subsubsection*{Detecting outlier perturbations}
We found the average response to each gene perturbation by grouping the perturbed dataset by target gene ID (using pandas, \url{http://pandas.pydata.org/}).
A Hotelling's $T^2$ test was used to determine whether the means of each gene knockdown was significantly different from the control population.
A robust covariance estimator was used to find the location and covariance of the control distribution (using scikit-learn).
Because the control distribution ($n=11,253$) is much larger than that for any particular gene ID ($n \approx 4$), the effect of the perturbed sample on the pooled covariance was neglected.

@@

\section*{Acknowledgments}
We thank Tsuyoshi Hirota, David Welsh, and John Abel for the helpful discussions, as well as the each of the authors of the original experiments for making their raw data available online.
% This work was supported by the National Institutes of Health/National Institute of General Medical Sciences under award number 1R01GM096873-01 and by the Institute for Collaborative Biotechnologies through grant W911NF-09-0001 from the U.S.\ Army Research Office.

>>>>

In the preceding sections, we have demonstrated that changes to single-cell intrinsic noise are sufficient to explain the observed changes in population-level damping.
However, experimental work has shown that cell-autonomous fibroblast cells have a distribution of mean free-running periods \cite{Leise2012}.
Indeed, prior to the availability of single-cell data, studies explored the possibility that differences in mean periods served as the mechanism behind population-level damping \cite{Izumo2003}.
While it is true that the dephasing of a group of oscillators can be caused by both variance in the mean period as well as cycle-to-cycle variability, intrinsic stochastic noise may play a more significant role. 
We show that there is greater variance in period on a cycle-to-cycle basis than on a cell-to-cell basis in cultured fibroblast cells (Figure~\ref{fig:cell2cell-variability}): individual cell period lengths have an average inner quartile range (IQR) of 3.18 hrs, while cell-to-cell average period has an IQR of 1.55 hrs. 
These results are independently confirmed by a previous study using Bayesian modeling, which found a standard deviation of periods within cells of 1.43 hrs, and 0.89 hrs across the population \cite{Cohen2012}.
A similar result has also been observed in dispersed SCN neurons, suggesting that while both intrinsic and extrinsic period heterogeneity likely contribute to the dephasing kinetics, cell-to-cell differences are less severe than cell-autonomous noise \cite{Herzog2004}.

It is also possible that damping rate changes due to siRNA or small molecule perturbation could be manifested through altering the system's extrinsic noise.
Such a change could be caused by an unequal uptake of siRNA or drug on a cell-to-cell basis, as has been demonstrated by a distribution of single-cell knockdown efficiency through flow cytometry \cite{Mousses2003}.
This effect would increase cell heterogeneity and lead to faster dephasing kinetics.
While differentiating between intrinsic and extrinsic noise sources from population-level data is possible in theory (Figure~\ref{fig:dephasing-rates}), these differences are likely not identifiable from typical population-level data (Figure~\ref{fig:computational_dephasing}).
This distinction would likely be possible with single-cell level data, as has been done in other experimental systems \cite{Elowitz2002}.
However, such an experiment would likely not be amenable to high-throughput methods.
Differences in damping rates are therefore best viewed as representative of changes to overall stochastic noise from both intrinsic and extrinsic factors.
Since both types of noise are important to determining the overall function of population-level rhythms, damping rates are still a valuable method of quantifying stochastic noise.

@@

% Overview of experiment - make it clear the insights from each portion. Use as changes for last reviewer.

In this study we have shown that the damping rate of population-level circadian oscillations can be changed by genetic or pharmacological perturbations.
As populations-level rhythms are determined by the coherence of many individual cells, desynchronization due to stochastic noise is a likely explanation for population-level damping.
Using single-cell data, we showed that population-level damping rate is proportional to single-cell noise.
Furthermore, we used a computational model to predict the changes in damping rate from two small molecules, demonstrating that changes to intrinsic stochastic noise are sufficient to explain the observed damping rate changes.

We have described a method by which changes to stochastic noise can be estimated from population-level circadian bioluminescence recordings.
An overview of the computational steps involved in our method are outlined in Figure~\ref{fig:comp_overview}.
While the method can be applied to existing experimental data, there are also practical considerations for the design of future experiments.
Because the damping rate must be inferred from the time-varying amplitude, collecting bioluminescence data for longer time periods yields more accurate results and reduces the potential impact of initial transient regions.
Additionally, a sampling rate that is high enough to confidently capture the peaks and troughs of gene expression is required -- in this study, the 2 hr sampling window of the siRNA screen proved sufficient.
While achieving such a rate is typically not difficult for bioluminescence experiments, it may limit the method's applicability in experiments where samples need to be analyzed at each time point.
We also note that there are many available tools for detrending and regressing time-series data.
While we prioritized computationally efficient methods (which could be scaled to genome-wide screens), the best methods for any particular application may vary depending on the data.

@@

\subsubsection*{Detecting outlier perturbations}
We found the average response to each gene perturbation by grouping the perturbed dataset by target gene ID (using pandas, \url{http://pandas.pydata.org/}).
A Hotelling's $T^2$ test was used to determine whether the means of each gene knockdown was significantly different from the control population.
While different siRNA constructs will have different knockdown efficiencies, grouping based on gene target helps to eliminate the effect of off-target activities. 
A robust covariance estimator was used to find the location and covariance of the control distribution (using scikit-learn).
Because the control distribution ($n=11,253$) is much larger than that for any particular gene ID ($n \approx 4$), the effect of the perturbed sample on the pooled covariance was neglected.

@@

\section*{Acknowledgments}
We thank Tsuyoshi Hirota, David Welsh, and John Abel for the helpful discussions, as well as the each of the authors of the original experiments for making their raw data available online.
We also thank the efforts of our four anonymous reviewers, who's suggestions have improved the quality of the manuscript.
% This work was supported by the National Institutes of Health/National Institute of General Medical Sciences under award number 1R01GM096873-01 and by the Institute for Collaborative Biotechnologies through grant W911NF-09-0001 from the U.S.\ Army Research Office.
